\documentclass[11pt,oneside,letterpaper]{article}

\usepackage{amsmath}
\usepackage{hyperref}
\usepackage[top=1in,bottom=1in,left=1in,right=1in]{geometry}
\usepackage{listings}
\usepackage{color}

\definecolor{dkgreen}{rgb}{0,0.6,0}
\definecolor{gray}{rgb}{0.5,0.5,0.5}
\definecolor{mauve}{rgb}{0.58,0,0.82}

\lstset{frame=tb,
    language=Java,
    aboveskip=3mm,
    belowskip=3mm,
    showstringspaces=false,
    columns=flexible,
    basicstyle={\small\ttfamily},
    numbers=left,
    numbersep=5pt,
    numberstyle=\tiny\color{gray},
    keywordstyle=\color{blue},
    commentstyle=\color{dkgreen},
    stringstyle=\color{mauve},
    breaklines=true,
    breakatwhitespace=true,
    tabsize=3
}

\begin{document}
\title{Applying Amdahl's Law to the Mandelbrot Fractal}
\author{Henry J Schmale}
\date{May 2015}
\maketitle

% Write the abstract
\begin{abstract}
The Mandelbrot fractal is an incredibly time consuming calculation.
The drawing of the Mandelbrot fractal can be speed up by multithreading.
However, there is a finite amount of performance that can be gained by
this method. The law governing this performance gain is called Amdahl's
Law.
\end{abstract}

\section{The Mandelbrot Set}
The Mandelbrot set is a set of numbers in the complex plane for which
a function remains bounded. The coloring of the fractal is determined
by the number of iterations it took to determine that it converges.
The formal definition is shown in equation \eqref{eq:mandeldef}.

\begin{equation}\label{eq:mandeldef}
    P_{c}: z \mapsto z^2 + c    
\end{equation}

Where in equation \eqref{eq:mandeldef}, $c$ is a complex parameter and the 
point in the grid currently being graphed. For each $c$, the behavior of the
sequence $(0, P_{c}(0), P_{c}(P_{c}(0)),...)$ is considered to determine
whether $P_{c}(z)$ escapes to infinity or remains bounded. The Mandelbrot
set is the set of all points $c$ that the sequence does not escape to
infinity.

\subsection{Drawing the Fractal}
The Mandelbrot fractal can be drawn using several methods. The
program being demonstrated today uses the escape time algorithm.
This method just brute forces every single point to determine what color
that pixel will be. This method is the simplest method to implement, but
it is also one of the slowest methods, because it is a brute force method.
However, it can be speed up with a periodicity check which is used in the
program being demonstrated.


% Next section on Amdahl's Law
\section{Amdahl's Law}
Amdahl's Law is used to find the maximum expected improvement to an overall
system when only one part of the system is improved. This law is most often
used in parallel computing to predict the theoretical maximum speedup when
using many processors to operate on a workload in parallel. This law is named
after the computer architect Gene Amdahl.

The speedup of a program using many processors in parallel is limited by the
amount of time needed for the serial fraction of the program. That is the
section that must be executed in a single thread of execution, or the part
that cannot be parallelized.

\subsection{Equations}
\label{sec:Amdahl-Equations}
Given:
\begin{itemize}
  \item $n \epsilon N$ The number of threads of execution.
  \item $B \epsilon [0,1]$ The fraction of the algorithm that is strictly serial.
\end{itemize}

$T(n)$ describes how long an algorithm will take to complete when running
with $n$ threads of execution, see equation \eqref{eq:amdahl_time}.

\begin{equation} \label{eq:amdahl_time}
    T(n) = T(1)(B + \frac{1}{n}(1-B))
\end{equation}

Therefore, the theoretical speedup $S(n)$ that can be had on a given algorithm
on a system capable of executing $n$ threads of execution is seen in equation
\eqref{eq:amdahl_speedup}
\begin{equation} \label{eq:amdahl_speedup}
    S(n) = \frac{T(1)}{T(n)}
         = \frac{T(1)}{T(1)(B+\frac{1}{n}(1-B))}
         = \frac{1}{B+\frac{1}{n}(1-B)}
\end{equation}

% Begin the next section that ties everything together
\section{Tying Amdahl's Law and the Mandelbrot Fractal Together}
The program I wrote to calculate the Mandelbrot fractal uses a nested loop to
draw to the screen, and that has a time complexity of at least $O(n^2)$. That
part is strictly serial. Since the drawing to the screen routine is strictly
serial, Amdahl's law applies. Now the worst case time complexity of calculating a
point in the Mandelbrot fractal using the escape time algorithm 
$O({colors} + 1)$, and the best case scenario is $O(1)$. This leads to a worst
case scenario for the whole program as $O(n^5)$.

Now plugging the numbers into Amdahl's equations yields the following speed up
when running on 5 threads (4 worker threads and 1 main thread). The speed up
yielded by multithreading can be seen in equation \eqref{eq:speedup5threads},
the final expected speed up running on 5 threads is 1.92 times faster than running
it in a single thread.
Given:
\begin{itemize}
    \item $T(1) = n^3$ The single threaded time complexity is $O(n^3)$
    \item $B = 2/5 = .4$ Roughly $2/5$ of the algorithm is serial
    \item The equations and givens in section \ref{sec:Amdahl-Equations}
        still apply.
\end{itemize}

\begin{equation} \label{eq:speedup5threads}
    \begin{align*}
        T(n)&= T(1)(B + \frac{1}{n}(1-B))\\
        T(5)&= (n^3)(.4 + \frac{1}{5}(1-.4))\\
        T(5)&= .52(n^3)\\
        S(5)&= \frac{T(1)}{T(5)} = \frac{n^3}{.52n^3} = 1.923
    \end{align*}
\end{equation}

% Distribute the source code in the handout
\section{The Source Code}
This section consists entirely of the source code to my multithreaded
Mandelbrot fractal program written in C++. However due to the use of
pthreads, it will only run under a Unix-like operating system. The 
entire contents of this project is available at 
\url{<https://github.com/HSchmale16/CalculusProject>}.
\lstinputlisting[language=C++]{mandelbrot.cpp}

\end{document}
